---
title: "New Antibody Lot Validation"
date: 
output: None
---

This document assess performance of new lots by assessing the raw data of 6 control 
cell lines for all antibodies. 

4 Metrics are used to assess performance:

 1) Correlation: between average of prior batches and new lot
 2) Technical replicate agreement (TRA): between all replicates
 3) General heatmap of prior samples and new with clustering
 4) RLE boxplot
 5) RLE heatmap

### Lot comparison
```{r, echo=TRUE}
oldlot="STP10004F"
newlot="STP10005F"
print(sprintf("oldlot = %s", oldlot))
print(sprintf("newlot = %s", oldlot))
#set to x dive QC directory
wd="/Volumes/Histopathology Shared Resource/CLINICAL/Nanostring/QC/Antibody Lot/STP10005F"
data_dir="/Volumes/Histopathology Shared Resource/CLINICAL/Nanostring/output"
```

Load Libraries
```{r libarary}
library(ruv)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(reshape2)
library(openxlsx)
library(gridExtra)
```

Load Metadata
```{r metadata for making eda metadata}
controls=c("MCF7","HCC1954","BT474","HeyA8","MDA468 control","MDA468+EGF")

md_file="/Users/patterja/Box/NANOSTRING/nanostring_metadata.xlsx"
md = read.xlsx(xlsxFile = md_file, sheet = "nanostring_metadata", check.names=F)
md$sampcolumn = paste0(md$Batch, "__", md$Sample.Name)

md$samp_deid = ifelse(make.names(as.character(md$Sample.Name)) %in% make.names(controls), yes=make.names(as.character(md$Sample.Name)), no="sample")
only_controls = md$Sample.Name %in% controls
control_samples = md$sampcolumn[md$Study %in% c("control")]

```

Batch sorting
```{r, echo=TRUE}
newlot_batch = "20190627_208420521020"
nbatch = "20190627"

md$batchdate = sapply(strsplit(md$Batch, split="_"), `[`, 1)
oldlot_batches = unique(md$Batch[md$batchdate < nbatch])
print("Old lot batches to summarize:")
print(oldlot_batches)
print("New lot:")
print(newlot_batch)
```

### Load raw data
```{r load raw data, echo=TRUE}

trawdat=data.frame()
batches = c()
for (batchid in oldlot_batches){
  rdfile = file.path(data_dir, batchid,"rawdata.txt")
  if (file.exists(rdfile)){
    dat = read.table(file = file.path(data_dir, batchid,"/rawdata.txt"), sep="\t", row.names=2, stringsAsFactors=F, header=T, check.names = T)
    dat[,c("CodeClass", "Accession")] <- NULL
    colnames(dat) = paste0(batchid, "__", colnames(dat))
    trawdat = rbind(trawdat, t(dat))
    batches = append(batches, batchid)
  }
}
oldlot_raw = t(trawdat)
dim(oldlot_raw)
colnames(oldlot_raw)

newlot_raw = read.table(file = file.path(data_dir, newlot_batch,"/rawdata.txt"), sep="\t", row.names=2, stringsAsFactors=F, header=T, check.names = T)
newlot_raw[,c("CodeClass", "Accession")] <- NULL

```

### Make mean of old lot data
```{r mean of old lots}


#old lot
ctls_oldlot = as.matrix(oldlot_raw[,make.names(colnames(oldlot_raw)) %in% make.names(control_samples)])
ctls_oldlot.m = melt(ctls_oldlot)
ctls_oldlot.m$batch = sapply(strsplit(as.character(ctls_oldlot.m$Var2), "__"), `[`, 1)
ctls_oldlot.m$ctlname = sapply(strsplit(as.character(ctls_oldlot.m$Var2), "__"), `[`, 2)
ctls_oldlot.m$ctl_probe = factor(paste0(ctls_oldlot.m$ctlname,"__", ctls_oldlot.m$Var1))

mean_ctls_oldlot = data.frame(value=tapply(ctls_oldlot.m$value, ctls_oldlot.m$ctl_probe, mean))
mean_ctls_oldlot$ctlname = sapply(strsplit(rownames(mean_ctls_oldlot), "__"), `[`, 1)
mean_ctls_oldlot$antibody = sapply(strsplit(rownames(mean_ctls_oldlot), "__"), `[`, 2)

#new lot
ctls_newlot = as.matrix(newlot_raw[,make.names(colnames(newlot_raw)) %in% make.names(controls)])
ctls_newlot.m = melt(ctls_newlot)
ctls_newlot.m$ctl_probe = factor(paste0(ctls_newlot.m$Var2,"__", ctls_newlot.m$Var1))


lotraw = data.frame(ctlprobe=rownames(mean_ctls_oldlot), 
                    mean_oldlot = mean_ctls_oldlot$value, 
                    newlot = ctls_newlot.m$value[match(rownames(mean_ctls_oldlot), as.character(ctls_newlot.m$ctl_probe))])

```

### Correlation

```{r}


#pairs plot for ERCC, average of old lot vs newlot
p_ercc = ggplot(lotraw[grepl("POS|NEG", lotraw$ctlprobe),]) +
  geom_point(aes(x=mean_oldlot, y=newlot)) +
  geom_abline(slope=1, intercept=0, color="red") +
  labs(title=paste0("Mean of old lots v. New lot=", newlot, 
                    "\nCorrelation of ERCC POS and NEG controls for each cell line+antibody",
                    "\nCorrelation R=", cor(lotraw[grepl("POS|NEG", lotraw$ctlprobe),"mean_oldlot"],lotraw[grepl("POS|NEG", lotraw$ctlprobe),"newlot"]))) +
    theme(plot.title = element_text(size=9))

p_ab = ggplot(lotraw[!grepl("POS|NEG", lotraw$ctlprobe),]) +
  geom_point(aes(x=mean_oldlot, y=newlot)) +
  geom_abline(slope=1, intercept=0, color="red") +
  labs(title=paste0("Mean of old lots v. New lot=", newlot, 
                    "\nCorrelation of Antibodies (no ERCC) for each cell line+antibody",
                    "\nCorrelation R=", cor(lotraw[!grepl("POS|NEG", lotraw$ctlprobe),"mean_oldlot"],lotraw[!grepl("POS|NEG", lotraw$ctlprobe),"newlot"]))) +
    theme(plot.title = element_text(size=9))

pdf(file = file.path(wd,"correlation.pdf"), width=6,height =8)
print(grid.arrange(p_ercc, p_ab, nrow=2))

for (b in unique(ctls_oldlot.m$batch)){
  blot = ctls_oldlot.m[ctls_oldlot.m$batch==b,]
  batchlot=data.frame(ctlprobe=blot$ctl_probe, 
                    blot = blot$value, 
                    newlot = ctls_newlot.m$value[match(blot$ctl_probe, as.character(ctls_newlot.m$ctl_probe))])
  
  colnames(batchlot)[which(colnames(batchlot)=="blot")]=b
  
  paste(b, "\n Lot = ", unique(md$Antibody.Lot[which(md$Batch==b)]))
  p_bercc = ggplot(batchlot[grepl("POS|NEG", batchlot$ctlprobe),]) +
    geom_point(aes_(x=as.name(b), y=quote(newlot))) +
    geom_abline(slope=1, intercept=0, color="red") +
    labs(title=paste0(b,"\nCorrelation of ERCC POS and NEG controls\n",
                     "Old lot=",unique(md$Antibody.Lot[which(md$Batch==b)]), " v. New lot= ", newlot,
                      "\nCorrelation R=", cor(batchlot[grepl("POS|NEG", batchlot$ctlprobe),c(2)],batchlot[grepl("POS|NEG", batchlot$ctlprobe),c(3)]))) +
    theme(plot.title = element_text(size=9))
  p_bab = ggplot(batchlot[!grepl("POS|NEG", batchlot$ctlprobe),]) +
    geom_point(aes_(x=as.name(b), y=quote(newlot))) +
    geom_abline(slope=1, intercept=0, color="red") +
    labs(title=paste0(b,"\nCorrelation of Antibodies (no ERCC)\n",
                     "Old lot=",unique(md$Antibody.Lot[which(md$Batch==b)]), " v. New lot= ", newlot,
                      "\nCorrelation R=", cor(batchlot[!grepl("POS|NEG", batchlot$ctlprobe),c(2)],batchlot[!grepl("POS|NEG", batchlot$ctlprobe),c(3)])))+
    theme(plot.title = element_text(size=9))
  print(grid.arrange(p_bercc, p_bab, nrow=2))

}
dev.off()
```

### Technical Repeatability Agreement

```{r}

mtraold = matrix(NA, nrow = 0, ncol = 4)
mtranew = matrix(NA, nrow = 0, ncol = 4)

for (ctrl in make.names(controls)){
  selctl.old = ctls_oldlot[!grepl("POS|NEG", rownames(ctls_oldlot)),grepl(ctrl, colnames(ctls_oldlot)),drop=F]
  selctl.new = ctls_newlot[!grepl("POS|NEG", rownames(ctls_newlot)),grepl(ctrl, colnames(ctls_newlot)),drop=F]
  #old lot tra
  tra_old = c()
    for (idx in 1:ncol(selctl.old)){
      temptra = log(as.matrix(selctl.old[,-idx])/as.numeric(selctl.old[,idx]))
      m=melt(as.matrix(temptra))
      m$Var2=colnames(selctl.old)[idx] #actual comparison is with this sample.
      tra_old = rbind(tra_old, m)
    }
  mtraold = rbind(mtraold, data.frame(ctrl,tra_old))
  tranew = log(as.matrix(selctl.old)/as.numeric(selctl.new))
  mn=melt(as.matrix(tranew))
  mn$Var2 = newlot_batch
  mtranew = rbind(mtranew, data.frame(ctrl, mn))
}

mtracomb = rbind(data.frame(mtraold,lot=oldlot), data.frame(mtranew, lot=newlot))
mtracomb$batch = sapply(strsplit(mtracomb$Var2, split="__"), `[`, 1)

pdf(file = file.path(wd,"tra_bycellline.pdf"), width=8,height =6)

for (ictrl in 1:length(unique(mtracomb$ctrl))){
  ctrl=as.character(unique(mtracomb$ctrl)[ictrl])
  ctrlmat = mtracomb[mtracomb$ctrl==ctrl,]
  ptra=ggplot(ctrlmat) + 
      geom_boxplot(aes(x=batch, y=value, fill=lot)) +
      facet_wrap(~ctrl, scale="free", ncol = 1) +
      geom_hline(yintercept =0, color="red",linetype="dashed") +
      labs(title=paste0("Distribution of TRA (technical replicate agreement)\nlog ratio of all replicates overall antibodies")) +
      theme(legend.position = "right",
            axis.text.x = element_text(size=9, colour=c("black"), angle=-90, hjust=0.5,vjust=0.5),
            plot.title = element_text(size=12),
            strip.text = element_text(size = 14))
  print(ptra)
}
dev.off()

pdf(file = file.path(wd,"tra_byantibody.pdf"), width=8,height =6)

for (iab in 1:length(unique(mtracomb$Var1))){
  ab=as.character(unique(mtracomb$Var1)[iab])
  mat = mtracomb[mtracomb$Var1==ab,]
  ptraab=ggplot(mat) + 
    geom_boxplot(aes(x=batch, y=value, fill=lot)) +
    facet_wrap(~Var1, scale="free", ncol = 2) +
    geom_hline(yintercept =0, color="red",linetype="dashed") +
    labs(title=paste0("Distribution of TRA (technical replicate agreement)\nlog",
    "ratio of all replicates overall celllines")) +
    theme(legend.position = "right",
          axis.text.x = element_text(size=9, colour=c("black"), angle=-90, hjust=0.5,vjust=0.5),
          plot.title = element_text(size=12),
          strip.text = element_text(size = 14))
  print(ptraab)
}

dev.off()

```

### General Heatmaps
```{r}

combctls = cbind(ctls_oldlot, ctls_newlot)
lcombctls = log(combctls+1)
mdcomb = data.frame(name=c(colnames(ctls_oldlot), paste0("newlot__", colnames(ctls_newlot))))
mdcomb$batch = sapply(strsplit(mdcomb$name, split="__"), `[`, 1)
mdcomb$cellline = sapply(strsplit(mdcomb$name, split="__"), `[`, 2)
rownames(mdcomb)=colnames(combctls)

pdf(file = file.path(wd,"signal_heatmap.pdf"), width=7,height =8)

pheat_ercc = pheatmap(mat = t(lcombctls[grepl("POS|NEG", rownames(ctls_oldlot)),]),
  color             = brewer.pal(n = 9, name = "YlOrRd"),
  border_color      = TRUE,
  show_colnames     = TRUE,
  show_rownames     = TRUE,
  annotation_row    = mdcomb[,-which(colnames(mdcomb)=="name")], 
  drop_levels       = TRUE,
  fontsize          = 8,
  fontsize_row      = 8,
  fontsize_col      = 8,
  main              = "log raw of ERCC", 
  cluster_rows      = T,
  cluster_cols      = F)

pheat_ab = pheatmap(mat = t(lcombctls[!grepl("POS|NEG", rownames(ctls_oldlot)),]),
  color             = brewer.pal(n = 9, name = "YlOrRd"),
  show_colnames     = TRUE,
  show_rownames     = TRUE,
  annotation_row    = mdcomb[,-which(colnames(mdcomb)=="name")], 
  border_color      = TRUE,
  drop_levels       = TRUE,
  fontsize          = 8,
  fontsize_row      = 8,
  fontsize_col      = 8,
  main              = "log raw of Antibodies", 
  cluster_rows      = T,
  cluster_cols      = T)

dev.off()
  
```

### RLE
```{r rle}
rleercc= rel_log_exp(Y=t(lcombctls[grepl("POS|NEG", rownames(ctls_oldlot)),]))
rleab= rel_log_exp(Y=t(lcombctls[!grepl("POS|NEG", rownames(ctls_oldlot)),]))

# RLE HEATMAP
pdf(file.path(wd,"RLE_heatmap.pdf"), width = 9.5, height = 8.5)
pheat_raw = pheatmap(mat = rleab,
                     color         = colorRampPalette(c("green", "black", "red"))(50),
                     border_color  = "darkgray",
                     show_colnames = TRUE,
                     show_rownames = TRUE,
                     annotation_row= mdcomb[,-which(colnames(mdcomb)=="name")], 
                     drop_levels   = TRUE,
                     fontsize      = 8,
                     fontsize_row  = 8,
                     fontsize_col  = 8,
                     main          = "Relative Log Expression of Antibodies", 
                     cluster_rows  = T,
                     cluster_cols  = T)


dev.off()
```
